%%%%%%%%%% Declaration Statement and Oat page %%%%%%%%%%%%%%%%%%%%%%
\section*{\centering Statement Lieu of Oath}
\flushleft
%\begin{doublespace}
I hereby declare that I have authored this Master’s Thesis independently, that I have not used other sources/resources than the ones declared, and that I have explicitly marked all material which has been quoted either literally or by content from the used sources. According to my knowledge, the content or parts of this Master’s Thesis has not been presented to any other examination authority and have not been published. I am aware that the respective work can be considered as a "fail" in the event of a false declaration.

\vspace*{12pt}
I understand that the Master’s Thesis will be scanned for plagiarism with the help of “Turnitin”, “Docoloc” or any other plagiarism detection software. By uploading the Master’s Thesis to LEA I automatically agree to its storage to the repository to enable comparison with future exams/papers submitted.

\vspace*{12pt}
This declaration does not grant the Hochschule Bonn-Rhein-Sieg permission to reproduce, publish or use this exam/paper without my consent. In any case, the examination and evaluation of my work must be carried out individually and independently from the results of the plagiarism detection service.

\vspace*{12pt}
\textit{Attention: Please fill in and sign the document, do not change the name of the file and upload it separately to LEA.  If this document is missing, your Master’s Thesis will not be reviewed and will therefore be considered as failed.}


\vspace*{4cm}
\rule{\textwidth}{0.4pt}\\
%\vspace*{pt}

\quad Date/Place	\hspace{6cm}   Name



\vspace*{3cm}

\rule{7cm}{0.4pt}\\
\quad Matriculation Number
%\end{doublespace}